Highly Available Transactions: Virtues and Limitations

### 2013-05-01 writeup/hat-appendix.tex
pass, plus spelling and grammar

In our non-sticky TA algorithm, clients buffer their writes until
commit, sending the final write to each data item to each respective
server. Along with each write, clients attach a the transaction ID and
a list of all data items written to in the transaction, $L_v$. Upon
receiving a new write to item $d$, a server places the write into a
\texttt{pending} queue and notifies all other servers responsible for
a data item in $L_v$ that it has received a write to $d$ with the
respective transaction ID. Once a server receives acknowledgments with
appropriate transaction ID for all items in $L_v$, it moves the write
from \texttt{pending} to the \texttt{stable} database (which applies
the writes according to highest-ID-wins). Accordingly, all writes that
should be made atomic with a write in \texttt{stable} are guaranteed
to be present on their respective servers, either in \texttt{pending}
or \texttt{stable}. Given the possibility of internal aborts, writes
to \texttt{pending} may require 2 RTTs: one to check constraints and
one to execute the above (but these checks \textit{cannot} cause other
transactions to block, as would be the case in a two-phase
\textit{commit} protocol).

For reads, each client tracks which values it should read. It
maintains a mapping (vector) from data items to transaction IDs. Upon
transaction start, the client clears its mapping, and, when it
performs a read from a data item, it sends the server its current ID
for the item. If the client does not supply an ID, the server returns
the latest value from \texttt{stable}. If the ID is non-empty, the
server either returns the matching value from \texttt{pending} or a
value with higher ID from \texttt{pending}---along with $L_v$ for the
write. When the client receives a response, it updates its local
vector with $L_v$. Accordingly, even if transaction's writes to two
keys $x$ and $y$ are present on the replica for $x$'s \texttt{stable}
and the replica for $y$'s \texttt{pending}, a client's reads from $x$
then $y$ will return a transactionally atomic pair of values because
the replica for $y$ will know to check \texttt{pending}.

>>>>

In this section, we provide additional information on our two-phase
Transactional Atomicity algorithm.

The below pseudocode describes a straightforward implementation of the
algorithm from Section~\ref{sec:ta}. Replicas maintain two sets of
data items: \texttt{pending} (which contains writes that are not yet
pending stable) and \texttt{good} (which contains writes that are
pending stable). Incoming writes are added to \texttt{pending} and
moved to \texttt{good} once they are stable. Stable pending
calculation is performed by counting the number of acknowledgments
(\texttt{acks}) of writes from other replicas, signifying their
receipt of the transaction's writes in their
own \texttt{pending}. Accordingly, the algorithm maintains the
invariant that the transactional siblings (or suitable replacements)
for every write in \texttt{good} are present on their respective
replicas. Clients use the RC algorithm from
Section~\ref{sec:isolation} and keep a map of data items to timestamps
(\texttt{required})---effectively a vector clock whose entries are
data items (not processes, as is standard).

\vspace{1em}\noindent\textbf{Example.} To illustrate this algorithm consider an execution of the following transactions:
\begin{align*}
\small\vspace{-1em}
T_1 &: w_x(1)~w_y(1)
\\T_2 &: r_x(1)~r_y(a)
\end{align*}
For TA, the system must ensure that $a=1$. Consider a system with one
replica for each of $x$ and $y$ (denoted $R_x$ and $R_y$). Say client
$c_1$ executes $T_1$ and commits. The client chooses timestamp $10001$
(whereby the last $4$ digits are reserved for client ID to ensure
uniqueness) and sends values to $w_x(1)$ to $R_x$ and $R_y$ with
$tx\_keys = \{x,y\}$ and $timestamp=10001$. $R_x$ and $R_y$ place
their respective writes into their individual \texttt{pending} sets
and send acknowledgment for transaction timestamped $10001$ to the
other. Consider the case where $r_x$ has seen $r_y$'s acknowledgment
but not vice versa: $w_x(1)$ will be in \texttt{good} on $R_x$
but \texttt{pending} on $R_y$. Now, if another client $c_2$ executes
$T_2$ and reads $x$ from $R_x$, it will read $x=1$. $c_2$ will update
its \texttt{required} vector so that $\texttt{required}=\{x:10001,
y:10001\}$. When $c_2$ reads $y$ from $R_y$, it will specify
$ts=10001$ and $R_y$ will not find a suitable write
in \texttt{good}. $R_y$ will instead return $y=1$
from \texttt{pending}.

\vspace{1em}\noindent\textbf{Overheads.} TA requires overhead both on client and server. First, clients must store a vector (\texttt{required}) for the duration of a transaction in order to determine what values to read. Second, each write contains metadata ($tx\_keys$) linear in the number of writes in the transaction. (Note that we could have stored a different timestamp with every write, but a single timestamp suffices and is more efficient) These overheads are modest for short transaction lengths but they increase value sizes on disk, in memory, and over the network. Third, on the server side, each client \textsc{put} incurs two back-ends \texttt{put}s: one into \texttt{pending} and \texttt{good}. If keeping a single value for \texttt{good}, the second \texttt{put} into \texttt{good} (as is the case with normal, single-valued eventually consistent operation) is really a \texttt{get} then a possible \texttt{put}: the server first checks if a higher-timestamped value has been written and, if not, it overwrites the current value. The \texttt{put} into \texttt{pending} is simpler: if message delivery is exactly-once, then \texttt{put}s can be made unconditionally.

\vspace{1em}\noindent\textbf{Optimizations.} The use of \texttt{pending} and \texttt{required} is effectively to
handle the above race condition, where some but not all replicas have
realized that a write is pending stable. With this in mind, there are
several optimizations we can perform:

If a replica serves a read from \texttt{pending}, it must mean one of
two things: either the write is stable pending or the writing client
for the write is requesting the data item (this second case does not
occur in the below pseudocode since own-writes are served out
of \texttt{write\_buffer}). In the former case, the replica can safely
install the write into \texttt{good}.

As we mention in Section~\ref{sec:ta}, the size of \texttt{good} can
be limited to a single item. Recall that any item can be served
from \texttt{good} as long as its timestamp is greater than the
required timestamp. Accordingly, if replicas store multiple writes
in \texttt{good}, any time they can serve a write from \texttt{good},
the highest-timestamped write is always a valid response candidate
(and, indeed, returning the highest-timestamped write provides the
best visibility). Note, however, that serving a higher-than-requested
timestamp from \texttt{pending} is dangerous: doing so might mean
returning a write that is not pending stable and, via appropriate
unfortunate partition behavior, force a client to stall while an
available replica waits for the write to arrive in \texttt{pending}.

The below pseudocode implements a push-based acknowledgment scheme: replicas
eagerly notify other replicas about \texttt{pending} writes. This need
not be the case and, indeed, it is wasteful: if a write
in \texttt{pending} has lower timestamp than the highest timestamped
write in \texttt{good}, it can safely be discarded. Especially for
data items with high temporal locality or data stores that are
replicated over WAN, this \texttt{pending} invalidation may be a
frequent event. Instead, replicas can periodically poll other replicas
for acknowledgments (i.e., asking for either a match
in \texttt{pending} or a higher timestamp in \texttt{good}); this
reduces both network utilization and, in the case of
durable \texttt{pending} and \texttt{good} sets, disk utilization.

There are ample opportunities for batching of pending and good
acknowledgments. Batching is easier in a push-based acknowledgment
scheme, as pull-based acknowledgments may require invalidation to
actually reduce network transfers. Moreover, if a replica is
responsible for several writes within a transaction, \textrm{notify}
messages can be coalesced. We have also observed that batching
multiple background operations together (particularly for
anti-entropy) allows for more efficient compression---especially
important in network-constrained environments.

\vspace{1em}\noindent\textbf{Implementation note.} Our implementation from Section~\ref{sec:evaluation} implements the below pseudocode, with batching between \textrm{notify} operations and a single-item \texttt{good}. We have not yet investigated the efficiency of a pull-based approach (mostly due to our choice of RPC library, which unfortunately does not currently handle multiple outstanding requests per socket). We have also not experimentally determined the visibility effect of the side-channel pending stable notifications as in the first optimization.

\vspace{1em}\noindent\textbf{Algorithmic improvements.} We are actively investigating more efficient alternatives to the algorithm presented here. In particular, a lower-overhead (but still highly available) implementation that used less metadata would be appealing. A system using server-side vector clocks is feasible but does not necessarily reduce overhead because the number of items written per transaction is likely smaller than the number of servers in a system. Similarly, we have observed an algorithm trade-off between the size of $ts_{required}$ and write metadata. While it is certainly possible to reduce overheads by reducing availability~\cite{eiger}, we believe high availability is a worthwhile goal for future exploration.

### 2013-05-01 writeup/hat-vldb.tex
pass, plus spelling and grammar

\documentclass{vldb}
\let\proof\relax
\let\endproof\relax
\usepackage{draftwatermarkbottom, draftwatermarktop, scrextend, MnSymbol}

@@

\begin{abstract}
\noindent To minimize network latency and remain online during server
failures and network partitions, many modern distributed data storage
systems eschew transactional functionality, which provides strong
semantic guarantees for groups of multiple operations over multiple
data items. In this work, we consider the problem of providing Highly
Available Transactions (HATs): transactional guarantees that do not
suffer unavailability during system partitions or incur high network
latency.  We introduce a new taxonomy of highly available systems and
analyze existing ACID and distributed consistency guarantees to
identify which can and cannot be achieved in HAT systems. This unifies
the literature on weak isolation, data consistency, and highly
available systems. We analytically and experimentally quantify the
availability and performance benefits of HATs---two to three orders of
magnitude over wide-area networks---as well as their necessary
semantic compromises.
\end{abstract}\vspace{-.5em}

@@

%APPENDIX is optional.
% ****************** APPENDIX **************************************
% Example of an appendix; typically would start on a new page
%\pagebreak
%
\pagebreak
\begin{appendix}
%\input{availability-appendix.tex}

\input{hat-appendix.tex}
\end{appendix}
\end{document}

>>>>

\documentclass{vldb}
\let\proof\relax
\let\endproof\relax
\usepackage{draftwatermarkbottom, draftwatermarktop}
\usepackage{scrextend, MnSymbol}

@@

\begin{abstract}
\noindent To minimize network latency and remain online during server
failures and network partitions, many modern distributed data storage
systems eschew transactional functionality, which provides strong
semantic guarantees for groups of multiple operations over multiple
data items. In this work, we consider the problem of providing Highly
Available Transactions (HATs): transactional guarantees that do not
suffer unavailability during system partitions or incur high network
latency.  We introduce a new taxonomy of highly available systems and
analyze existing ACID and distributed consistency guarantees to
identify which can and cannot be achieved in HAT systems. This unifies
the literature on weak isolation, replica consistency, and highly
available systems. We analytically and experimentally quantify the
availability and performance benefits of HATs---often two to three
orders of magnitude over wide-area networks---as well as their
necessary semantic compromises.
\end{abstract}\vspace{-.5em}

@@

%APPENDIX is optional.
% ****************** APPENDIX **************************************
% Example of an appendix; typically would start on a new page
%\pagebreak
%
\pagebreak
%\begin{appendix}
%\input{availability-appendix.tex}

\input{hat-appendix.tex}
%\end{appendix}
\end{document}

### 2013-05-01 writeup/intro.tex
pass, plus spelling and grammar

The last decade has seen a significant shift in the design of
large-scale database backends, from the use of transactional
RDBMSs~\cite{bernstein-book, gray-isolation, gray-virtues} to the
widespread adoption of loosely-consistent distributed key-value
stores~\cite{bigtable, pnuts, dynamo}. The 2000 introduction of
Brewer's CAP Theorem, which stated that a highly available system
cannot provide strong consistency guarantees in the presence of
network partitions, helped spur this shift~\cite{brewer-slides}. As
formally proven~\cite{gilbert-cap}, the CAP Theorem concerns
linearizability: the ability to read the most recent write to a data
item that is replicated across servers~\cite{herlihy-art}. Yet the
implications of CAP---which concern a fairly narrow form of
distributed consistency---have often been conflated with the ability
to provide ACID database properties~\cite{brewer-slides, hn,
  foundation-article}; this misunderstanding has led to substantial
confusion regarding replica consistency, transactional isolation, and
high availability. The recent resurgence of transactional systems like
Spanner~\cite{spanner} suggests that programmers value these
semantics, but, unfortunately, existing transactional data stores do
not provide availability in the presence of
partitions~\cite{middleware-db, foundation-article,
  hstore,generalizedsnapshot, mdcc, krikellas-bargain, eiger, walter,
  calvin}.

Indeed, serializable transactions---the gold standard of traditional
ACID databases---are not achievable with high availability in the
presence of partitions~\cite{davidson-survey}. However, database
systems have a long tradition of providing weaker isolation and
consistency guarantees~\cite{adya, ansicritique, gray-virtues,
  gray-isolation, kemme-thesis}. Today's ACID and NewSQL databases
often employ weak isolation models due to their concurrency and
performance benefits; weak isolation is overwhelmingly the default
setting in these stores and is often the only option offered
(Section~\ref{sec:modernacid}). While weak isolation levels do not
provide serializability for general-purpose transactions, they are
apparently strong enough to deliver acceptable behavior to many
application programmers and are substantially stronger than the
semantics provided by current highly available systems. This raises a
natural question: which of these semantics can be provided with high
availability?

Unfortunately, the relationship between ACID semantics and high
availability has not been well-explored. We have a strong
understanding of weak isolation in the single-server context from
which it originated~\cite{adya, ansicritique, gray-isolation} and many
papers offer techniques for providing distributed
serializability~\cite{bernstein-book, spanner, daudjee-session,
  hstore, krikellas-bargain, calvin, kemme-classification} or snapshot
isolation~\cite{daudjee-snapshot,generalizedsnapshot, kemme-snapshot,
  walter}. Additionally, the distributed computing and parallel
hardware literature contains many consistency models for single
operations on replicated objects~\cite{pnuts, herlihy-art, eiger, cac,
  sessionguarantees}. However, the literature lends few clues for
providing semantic guarantees for multiple operations operating on
multiple data items in a highly available distributed environment.

Our main contributions in this paper are as follows. We fill the gap
between the plethora of previously proposed isolation and consistency
models and the goal of high availability.  We classify which among the
wide array of models are achievable with high availability, denoting
them as {\em Highly Available Transactions} (HATs). In doing so, we
demonstrate that although many implementations of HAT semantics are
not highly available, this is an artifact of the implementations
rather than an inherent property of the semantics. Our investigation
shows that, besides serializability, Snapshot Isolation and Repeatable
Read isolation are not HAT-compliant, while most other isolation
levels are achievable with high availability. We also demonstrate that
many weak data consistency models from distributed systems are also
both HAT-compliant and simultaneously achievable with traditional ACID
properties.

Our investigation is based on both impossibility results and several
constructive, proof-of-concept algorithms. For example, Snapshot
Isolation and Repeatable Read isolation are not HAT-compliant because they
require detecting conflicts between concurrent updates (as needed for
preventing Lost Updates or Write Skew phenomena), which we prove is
unavailable. However, Read Committed isolation, Transactional
Atomicity, and many of the weaker data consistency models from
database and distributed systems are achievable via novel algorithms
that rely on multi-versioning and limited client-side caching. For
several guarantees, such as causal consistency with phantom prevention
and ANSI Repeatable Read, we consider a modified form of high
availability in which clients ``stick to'' (i.e., have affinity with)
at least one server (a property which is often implicit in the
distributed systems literature~\cite{herlihy-art, eiger, cac} but
which requires explicit consideration in a client-server replicated
database context). This sticky availability is widely
employed~\cite{eiger, vogels-defs} but is less restrictive (and
therefore more easily achievable) than traditional high availability.

Our results demonstrate that a wide range of semantic guarantees can
be delivered with availability in face of partitions, but which
guarantees are worthwhile in practice? We study the virtues and
limitations of both classic and HAT systems by surveying practitioner
accounts and research literature, performing experimental analysis on
modern cloud infrastructure, and analyzing representative applications
for their semantic requirements. Our experiences with a HAT prototype
running across multiple geo-replicated datacenters indicate that HATs
offer a one to three order of magnitude latency decrease compared to
traditional distributed serializability protocols, and they can
provide acceptable semantics for a wide range of programs, especially
those with monotonic logic and commutative updates~\cite{calm, crdt}. HAT systems can also enforce arbitrary foreign key
constraints on multi-item updates and sometimes provide limited
uniqueness guarantees. However, particularly for programs with
non-monotonic logic, HATs can fall short, requiring
sometimes-unavailable mechanisms that are more
coordination-intensive. Overall, this paper quantifies the benefits of
Highly Available Transactions can offer as well as the necessary
semantic restrictions they place on applications.

>>>>

The last decade has seen a significant shift in the design of
large-scale database systems, from the use of transactional
RDBMSs~\cite{bernstein-book, gray-isolation, gray-virtues} to the
widespread adoption of loosely consistent distributed key-value
stores~\cite{bigtable, pnuts, dynamo}. The 2000 introduction of
Brewer's CAP Theorem, which stated that a highly available system
cannot provide strong consistency guarantees in the presence of
network partitions, helped spur this shift~\cite{brewer-slides}. As
formally proven~\cite{gilbert-cap}, the CAP Theorem concerns
linearizability: the ability to read the most recent write to a data
item that is replicated across servers~\cite{herlihy-art}. Yet the
implications of CAP---which concern a fairly narrow form of
distributed consistency---have often been conflated with the ability
to provide ACID database properties~\cite{hn,brewer-slides,
  foundation-article}; this misunderstanding has led to substantial
confusion regarding replica consistency, transactional isolation, and
high availability. The recent resurgence of transactional systems like
Spanner~\cite{spanner} suggests that programmers value these
semantics, but, unfortunately, existing transactional data stores do
not provide availability in the presence of
partitions~\cite{middleware-db, foundation-article,
  hstore,spanner,generalizedsnapshot, mdcc, krikellas-bargain, eiger,
  walter, calvin}.

Indeed, serializable transactions---the gold standard of traditional
ACID databases---are not achievable with high availability in the
presence of partitions~\cite{davidson-survey}. However, database
systems have a long tradition of providing weaker isolation and
consistency guarantees~\cite{adya, ansicritique, gray-virtues,
  gray-isolation, kemme-thesis}. Today's ACID and NewSQL databases
often employ weak isolation models due to concurrency and performance
benefits; weak isolation is overwhelmingly the default setting in
these stores and is often the only option offered
(Section~\ref{sec:modernacid}). While weak isolation levels do not
provide serializability for general-purpose transactions, they are
apparently strong enough to deliver acceptable behavior to many
application programmers and are substantially stronger than the
semantics provided by current highly available systems. This raises a
natural question: which of these semantics can be provided with high
availability?

To date, the relationship between ACID semantics and high
availability has not been well explored. We have a strong
understanding of weak isolation in the single-server context from
which it originated~\cite{adya, ansicritique, gray-isolation} and many
papers offer techniques for providing distributed
serializability~\cite{bernstein-book, spanner, daudjee-session,
  hstore, krikellas-bargain, calvin, kemme-classification} or snapshot
isolation~\cite{daudjee-snapshot,generalizedsnapshot, kemme-snapshot,
  walter}. Additionally, the distributed computing and parallel
hardware literature contains many consistency models for single
operations on replicated objects~\cite{pnuts, herlihy-art, eiger, cac,
  sessionguarantees}. However, the literature lends few clues for
providing semantic guarantees for multiple operations operating on
multiple data items in a highly available distributed environment.

Our main contributions in this paper are as follows. We fill the gap
between the many previously proposed isolation and consistency models
and the goal of high availability.  We classify which among the wide
array of models are achievable with high availability, denoting them
as {\em Highly Available Transactions} (HATs). In doing so, we
demonstrate that although many implementations of HAT semantics are
not highly available, this is an artifact of the implementations
rather than an inherent property of the semantics. Our investigation
shows that, besides serializability, Snapshot Isolation and Repeatable
Read isolation are not HAT-compliant, while most other isolation
levels are achievable with high availability. We also demonstrate that
many weak replica consistency models from distributed systems are both
HAT-compliant and simultaneously achievable with several ACID
properties.

Our investigation is based on both impossibility results and several
constructive, proof-of-concept algorithms. For example, Snapshot
Isolation and Repeatable Read isolation are not HAT-compliant because
they require detecting conflicts between concurrent updates (as needed
for preventing Lost Updates or Write Skew phenomena), which we show
is unavailable. However, Read Committed isolation, Transactional
Atomicity, and many of the weaker data consistency models from
database and distributed systems are achievable via algorithms that
rely on multi-versioning and limited client-side caching. For several
guarantees, such as causal consistency with phantom prevention and
ANSI Repeatable Read, we consider a modified form of high availability
in which clients ``stick to'' (i.e., have affinity with) at least one
server---a property which is often implicit in the distributed systems
literature~\cite{herlihy-art, eiger, cac} but which requires explicit
consideration in a client-server replicated database context. This
sticky availability is widely employed~\cite{eiger, vogels-defs} but
is less restrictive (and therefore more easily achievable) than
traditional high availability.

Our results demonstrate that a wide range of semantic guarantees can
be delivered with availability in face of partitions, but which
guarantees are worthwhile in practice? We study the virtues and
limitations of both classic and HAT systems by surveying practitioner
accounts and research literature, performing experimental analysis on
modern cloud infrastructure, and analyzing representative applications
for their semantic requirements. Our experiences with a HAT prototype
running across multiple geo-replicated datacenters indicate that HATs
offer a one to three order of magnitude latency decrease compared to
traditional distributed serializability protocols, and they can
provide acceptable semantics for a wide range of programs, especially
those with monotonic logic and commutative updates~\cite{calm,
  crdt}. HAT systems can also enforce arbitrary foreign key
constraints for multi-item updates and, in some cases, provide limited
uniqueness guarantees. However, particularly for programs with
non-monotonic logic, HATs can fall short, requiring
sometimes-unavailable mechanisms that are more coordination
intensive. Overall, this paper quantifies the benefits that Highly
Available Transactions offer as well as the necessary restrictions
that they place on applications.

### 2013-05-01 writeup/model.tex
pass, plus spelling and grammar

To understand which guarantees can be provided with high availabilty,
we must first define what high availability means. In this section, we
will formulate a model that captures a range of availability models,
including ``true'' high availability, availability with stickiness,
and transactional availability.

Informally, highly available algorithms ensure ``always on'' operation
and guaranteed low latency. If users of a highly available system are
able to contact a (set of) server(s) in a system, they are guaranteed
a response; this means servers will not need to communicate with any
others. If servers are partitioned from one another, they do not need
to stall in order to provide clients a ``safe'' response to
operations. This lack of fast-path coordination also means that a
highly available system also provides low latency~\cite{abadi-pacelc};
in a wide-area setting, this means that clients need not wait for
cross-datacenter communication. To properly describe whether a
\textit{transactional} system is highly available, we need to describe
what servers a client must contact as well as what kinds of responses
a server can provide, especially given the possibility of aborts.

@@

Traditionally, a system provides {\em high availability} if every user
that can contact a correct (non-failing) server eventually receives a
response from that server, even in the presence of arbitrary,
indefinitely long network partitions between
servers~\cite{gilbert-cap}.\footnote{Under this definition, systems
  that require a majority of servers to be online or that require the
  database to function correctly with high (but not perfect)
  probability are not available.} As in a standard distributed
database, designated servers might perform operations for different
data items; a server that can handle an operation for a given data
item is called a \textit{replica} for that item.\footnote{There is a
  further distinction between a \textit{fully replicated} system, in
  which all servers are replicas for all data items, and a
  \textit{partially replicated} system, in which at least one server
  acts as a replica for only a subset of all data items. For
  generality, and, for applicability to many modern ``sharded'' or
  ``partitioned'' data storage systems~\cite{dynamo, pnuts, bigtable,
    spanner, hstore}, we consider partially replicated systems.}

In addition to high availability, which allows operations on any replica,
distributed algorithms often assume a model in which clients always
contact the same (sets of) replica(s) across subsequent operations. As
we will discuss in Section~\ref{sec:hats}, clients can ensure
continuity between operations (e.g., reading their prior updates to a
data item) by maintaining affinity or ``stickiness'' with a server or
set of servers~\cite{vogels-defs}. Simultaneously maintaining
availability and stickiness requires fate-sharing between the client
and its server: if a client's sticky servers fail or if the client is
partitioned from its sticky servers, then it may face
unavailability. We say that a system provides \textit{sticky
  availability} if, whenever a client's operation is executed against
a correct server that has observed all of its prior operations, it
eventually receives a response, even in the presence of indefinitely
long partitions. A client may choose to become sticky available by
acting as a server itself; for example, a client might cache its reads
and writes~\cite{bolton, sessionguarantees, swift}. Any guarantee
achievable in a highly available system is achievable in a sticky high
availability system but not vice-versa.

>>>>

To understand which guarantees can be provided with high availability,
we must first define what high availability means. In this section, we
will formulate a model that captures a range of availability models,
including high availability, availability with stickiness, and
transactional availability.

Informally, highly available algorithms ensure ``always on'' operation
and guaranteed low latency. If users of a highly available system are
able to contact a (set of) server(s) in a system, they are guaranteed
a response; this means servers will not need to communicate with any
others. If servers are partitioned from one another, they do not need
to stall in order to provide clients a ``safe'' response to
operations. This lack of fast-path coordination also means that a
highly available system also provides low latency~\cite{abadi-pacelc};
in a wide-area setting, clients of a highly available system need not
wait for cross-datacenter communication. To properly describe whether
a \textit{transactional} system is highly available, we need to
describe what servers a client must contact as well as what kinds of
responses a server can provide, especially given the possibility of
aborts.

@@

Traditionally, a system provides {\em high availability} if every user
that can contact a correct (non-failing) server eventually receives a
response from that server, even in the presence of arbitrary,
indefinitely long network partitions between
servers~\cite{gilbert-cap}.\footnote{Under this definition, systems
  that require a majority of servers to be online is not
  available. Similarly, a system which ensures that clients receive a
  response with high (but not perfect) probability is not available.}
As in a standard distributed database, designated servers might
perform operations for different data items; a server that can handle
an operation for a given data item is called a \textit{replica} for
that item.\footnote{There is a further distinction between a
  \textit{fully replicated} system, in which all servers are replicas
  for all data items, and a \textit{partially replicated} system, in
  which at least one server acts as a replica for only a subset of all
  data items. For generality, and, for applicability to many modern
  ``sharded'' or ``partitioned'' data storage systems~\cite{
   bigtable, pnuts,  spanner, dynamo, hstore}, we consider partially
  replicated systems.}

In addition to high availability, which allows operations on any replica,
distributed algorithms often assume a model in which clients always
contact the same (sets of) replica(s) across subsequent operations. As
we will discuss in Section~\ref{sec:hats}, clients can ensure
continuity between operations (e.g., reading their prior updates to a
data item) by maintaining affinity or ``stickiness'' with a server or
set of servers~\cite{vogels-defs}. Simultaneously maintaining
availability and stickiness requires fate sharing between the client
and its server: if a client's sticky servers fail or if the client is
partitioned from its sticky servers, then it may face
unavailability. We say that a system provides \textit{sticky
  availability} if, whenever a client's operation is executed against
a correct server that has observed all of its prior operations, it
eventually receives a response, even in the presence of indefinitely
long partitions. A client may choose to become sticky available by
acting as a server itself; for example, a client might cache its reads
and writes~\cite{bolton, sessionguarantees, swift}. Any guarantee
achievable in a highly available system is achievable in a sticky high
availability system but not vice-versa.
